### 2017-01-16 thesis/methods/methods.tex
tmp commit

The samples were grouped based on the sample BMI status (obese or non-obese group), and a linear model was fitted to the data based on the sample grouping.
Empirical Bayes statistics was used to identify the differentially expressed genes between the two groups, and the top scoring genes were listed.
The error rate of the genes were \gls{fdr} controlled for multiple hypothesis testing.

>>>>

For gene expression analysis, or differential expression analysis, the \textit{limma} package in \gls{r} was used \citep{Ritchie2015}.
Since there were thousands of genes to be hypothesis tested, adjustment for multiple hypothesis testing had to be considered in order to identify the truly \glspl{deg}.

\subsection{Limma}
\label{sub:limma}

\textit{limma} is a package that contains variety of tools to analyse microarray data using linear models-based methods, developed by \citet{Ritchie2015}.
For gene expression analysis, \textit{limma} package fits a linear model in a gene-wise manner and produce test statistics that allow the assessment of whether the gene is differentially expressed or not \citep{Ritchie2015}.

Before the data was analysed using the \textit{limma} analysis pipeline, it was normalised as described in \cref{sub:data_normalisation}: \nameref{sub:data_normalisation}.
In addition to this, a design matrix that describes the experimental design was created from the clinical data.
To create the design matrix, the samples were divided into two groups, the obese group and the non-obese group, and the constructed group information was used in the \texttt{model.matrix} function (in \textit{limma} package) to form the design matrix.

\texttt{lmFit} function (\textit{limma}) was used to fit a linear model to the normalised data, using the experiental design information from the design matrix.
The output of this function was used in the \texttt{eBayes} function (\textit{limma}) to identify the \glspl{deg} from the data.
In \texttt{eBayes} function, statistical parameters are estimated from the data and these parameters are used in the empirical Bayesian approach to calculate the summary statistics used for the ranking and identification of \glspl{deg} \citep{Smyth2004}.

The summary statistics from the \texttt{eBayes} function can be displayed with the \texttt{topTable} function (\textit{limma}) and includes: the estimate of the fold change of the gene expression in log$_2$, average gene expression level in log$_2$, moderated \textit{t}-statistic, raw p-value of the gene, multiple hypothesis testing adjusted p-value, and \textit{B}-statistic.
The first value shows the estimate of the log$_2$ fold change of the gene expression compared with the reference group, so this represents the log$_2$ fold change of the gene expression in the obese samples relative to the non-obese group.
Second value presents the log$_2$ value of the average expression of the gene across all of the arrays/samples.
Moderated \textit{t}-statistic is the same as normal \textit{t}-statistic, but its standard error has been adjusted by a simple Bayesian method \citep{Smyth2005}.
Raw p-value and adjusted p-value represent the p-value of the gene before and after it has been corrected for multiple hypothesis testing, respectively.
Lastly, the value of \textit{B}-statistic represents the log-odds that the gene is differentially expressed, where \textit{B}-statistic of 0 shows that there is a 50\% chance the gene is differentially expressed \citep{Smyth2005}.

From these summary statistics, the most likely \glspl{deg} were chosen from the list of significant genes by setting the threshold of the p-value to be either less than 1\% or 5\%.
In the case where there were more than 1000 \glspl{deg}, the top 799 genes were picked from the list, as this was the number of genes found in the \citet{Creighton2012} study; otherwise, as many significant genes identified were taken from the list.

\subsection{Multiple hypothesis testing correction}
\label{sub:multiple_hypothesis_testing_correction}

In an experiment where there are multiple hypotheses being tested, the rate or proportion of \gls{type1} appearing from the experiment must be controlled.
The probability of \gls{type1} occuring for a single hypopthesis is usually controlled at some significance threshold $\alpha$, which is usually set at 0.05 \citep{Shaffer1995}.
In a typical microarray experiment, there are over 20,000 gene probes to be tested for differential expression, and with a significance threshold of $\alpha = 0.05$, this would yield approximately 1,000 \glspl{type1}.
In other words, 1,000 gene probes would be identified as differentially expressed, when in fact they are not.

There are two broad classes of methods to correct for this problem: \gls{fwer} control and \gls{fdr} control.
With \gls{fwer} control, the method primarily aims to set the $\alpha$-value for each hypothesis testing ($\alpha_i$) such that the sum of all the $\alpha_i$ is equal to $\alpha$ \citep{Hochberg1987,Shaffer1995}.
Usually, $\alpha_i$ is set to $\frac{\alpha}{n}$, where $n$ is the number of hypothesis tests carried out in the experiment \citep{Shaffer1995}.
This highly conservative approach significantly improves the certainty of the result from the experiment, but at the same time it significantly increases the likelihood of missing the truly \glspl{deg}, or \glspl{type2}.

In contrast to the conservative \gls{fwer} control methods, the \gls{fdr} method developed by \citet{Benjamini1995a} control the \glspl{type1} while maintaining statistical \gls{power}.
The \gls{fdr} method controls the ``expected proportion of errors among the rejected hypotheses'' by adjusting the $\alpha$ to an alternative value, $q^*$, which is calculated from the p-values of the hypothesis tests \citep{Benjamini1995a}.
With \gls{fdr}, the p-values are ranked and ordered from the lowest to the highest p-value,

### 2017-01-16 thesis/misc/glossary.tex
Finished draft copy of introduction -- need to cut down on the contents and link everything together
Finished data normalisation draft
Finished batch correction section of methods
tmp commit

@@

\newacronym{bmi}{BMI}{Body Mass Index}

@@

\newacronym{ln}{LN}{Lymph node}
\newacronym{pr}{PR}{Progesterone receptor}
\newacronym{er}{ER}{Estrogen receptor}
\newacronym{her2}{HER2}{Human epidermal growth factor receptor 2}
\newacronym{svd}{SVD}{Singular value decomposition}
\newacronym{rma}{RMA}{Robust Multi Array}
\newacronym{mas}{MAS5}{Microarray Analysis Suite 5.0}
\newacronym{fdr}{FDR}{False Discovery Rate}

@@

@@

@@

@@

@@

>>>>

\newglossaryentry{type1}{
	name=Type I error,
	text=type I error,
	plural=type I errors,
	description={Incorrect rejection of a true null hypothesis; False positive}
}

\newglossaryentry{type2}{
	name=Type II error,
	text=type II error,
	plural=type II errors,
	description={Incorrect retaining of a false null hypothesis; False negative}
}

\newglossaryentry{power}{
	name=Power,
	text=power,
	description={Probability that it will reject a false null hypothesis; It is inversely related to the probability of making a type II error}
}

@@

\newacronym{bmi}{BMI}{body mass index}

@@

\newacronym{ln}{LN}{lymph node}
\newacronym{pr}{PR}{progesterone receptor}
\newacronym{er}{ER}{estrogen receptor}
\newacronym{her2}{HER2}{human epidermal growth factor receptor 2}
\newacronym{svd}{SVD}{singular value decomposition}
\newacronym{rma}{RMA}{robust multi array}
\newacronym{mas}{MAS5}{microarray analysis suite 5.0}
\newacronym{fdr}{FDR}{false discovery rate}
\newacronym{fwer}{FWER}{family-wise error rate}

@@

\newacronym{mtor}{mTOR}{mammalian target of rapamycin}

@@

\newacronym{deg}{DEG}{differentially expressed gene}

@@

\newacronym{pm}{PM}{perfect match}
\newacronym{mm}{MM}{mismatch}

@@

\newacronym{cpm}{cpm}{counts per million}
\newacronym{m}{$\mu$}{mean}
\newacronym{sd}{$\sigma$}{standard deviation}

@@

\newacronym{cdna}{cDNA}{complimentary DNA}
\newacronym{mrna}{mRNA}{messenger RNA}
\newacronym{ngs}{NGS}{next generation sequencing}
\newacronym{fc}{FC}{fold change}
